---
title: "I-SPY2 Radiogenomic Analysis"
subtitle: "Exploring Molecular and Radiological Features of Treatment Response in Breast Cancer"
author: "Nasim Bondar Sahebi"
date: "`r format(Sys.time(), '%Y-%m-%d')`"
output: 
  rmdformats::readthedown:
    toc_depth: 4
    theme: lumen
    highlight: espresso
    code_folding: show
    params:
      palette: "npg"
      size: 1
      showlabels: false
      ellipse: true
    css: "styles/custom.css"
    link-citations: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, results = 'asis')
knitr::opts_knit$set(root.dir = normalizePath(".."))
```

## Study Rationale & Analysis Framework

This analysis integrates multi-modal data from the I-SPY2 trial to investigate molecular and radiological features associated with neoadjuvant treatment response in breast cancer, following the radiogenomic framework proposed by [Sun et al. (PMID: 30120041)](https://pubmed.ncbi.nlm.nih.gov/30120041/).

### Data Modalities

- **Genomics:** Microarray expression profiling (Agilent 44K) for 987 patients
- **Radiomics:** MRI-derived tumor features from 985 patients using DCE-MRI across four timepoints (T0–T3)
- **Matched cohort:** 982 patients with both genomic and radiomic data

### Analysis Pipeline

**Phase 1: Genomics (Complete)**

1. Preprocess expression data, filter to 17,547 protein-coding genes
2. Differential expression analysis (R vs NR) — full cohort and per treatment
3. Pathway enrichment (GSEA + ORA) using Hallmark and KEGG gene sets
4. Gene–response correlation (DE hits, PredictIO signatures, immune/luminal markers)
5. Stratified signature analysis by treatment arm

**Key Genomic Findings:**

- **IO+Chemo (69 patients, 47 DE genes):**
  - Responders: immune-hot — LAMP3, CCL13, IDO1, CRABP1, CHI3L2, CXCL13, CXCL9, MARCO, UBD, KRT5
  - Non-responders: hormone-driven — CA12, AGR2, FOXA1, AR, ACOX2, GATA3, TFF1, TFF3, SCUBE2, MSMB
  - Pathways up in responders: interferon gamma, allograft rejection, interferon alpha, IL6/JAK/STAT3, E2F, G2M
  - Pathways up in non-responders: estrogen response early/late, EMT, myogenesis
- **Targeted (52 patients, 3 DE genes):**
  - Responders: GRB7, MNX1, PNMT (GRB7 is HER2-associated, co-amplified with ERBB2)
  - Pathways up in responders: mTORC1 signaling, interferon gamma, inflammatory response, glycolysis, hypoxia
  - Pathways up in non-responders: estrogen response early/late, WNT/beta-catenin, EMT
- **Chemo (179 patients):** 0 DE genes — no signal
- **Chemo+Targeted (687 patients):** 4,152 FDR significant genes but only 1 with high fold change: AGR3
- **Across all arms:** Estrogen-driven tumors resist treatment regardless of type

**Phase 2: Radiomics (In Progress)**

1. **T0 Feature Extraction:**
   - Extract baseline radiomic features using PyRadiomics (shape, first order, GLCM, etc.)
   - Include 3 features from Chitalia et al.(#TODO)
   - Filter small/missing segmentations
   - Select earliest AcquisitionDate per patient from H4H index CSV
2. **READII:**
   - Generate negative control images
   - Extract same feature sets on controls
3. **Delta-Radiomics (T1 − T0):**
   - Filter to patients with at least T0 and T1
   - Calculate feature differences (T1 − T0)
   - Visualize volume change
   - Identify features correlated with volume change
   - Identify features NOT correlated with volume → correlate with genomics
4. Correlate radiomic features with treatment response

**Phase 3: Radiogenomic Integration (Planned)**

Following [Sun et al. (PMID: 30120041)](https://pubmed.ncbi.nlm.nih.gov/30120041/), who linked CT radiomic features to CD8 cell infiltration using machine learning, we extend this approach to breast cancer MRI with multi-arm treatment stratification and longitudinal imaging.

1. **Correlation analysis:** radiomic features vs gene expression features, subset by feature class and treatment/receptor
2. **Unsupervised linking:** explore whether MRI features can distinguish immune-hot vs immune-cold tumors
3. **Predictive signatures:** build combined radiogenomic signature, test if it predicts treatment response
4. **Optional — FMCIB:** deep learning radiomics feature extraction

### Key Question

Can radiomics predict tumor shrinkage or immune phenotype better than genomics alone — or does combining both improve prediction?

---

## Overview

### Dataset Summary

- **Study Reference**: [PMID: 35623341](https://pubmed.ncbi.nlm.nih.gov/35623341/)  
- **Clinical Trial**: [ISPY2/NCT01042379](https://clinicaltrials.gov/study/NCT01042379)  
- **Data Source**: [GEO: GSE194040](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE194040)  
- **Study Population**:  
  - **Genomics Data**: 987 patients (mRNA & RPPA profiling)  
  - **Radiomics Data**: 985 patients (MRI scans)  
- **Platforms Used**:
  - **Genomics**: Agilent 44K arrays  
  - **Radiomics**: MRI (DCE-MRI)   

#### Treatment Details

- **Total Patients (Microarray)**: 987  
- **Total Patients (MRI)**: 985 (719 I-SPY2 cases, 266 ACRIN-6698 cases)  
- **Treatment Groups**:
  - **IO + Chemo**: 69 patients  
  - **Chemo + Targeted Therapy**: 687 patients  
  - **Chemo Alone**: 179 patients  
  - **Targeted Therapy Alone**: 52 patients  
- **Key Treatment Arms**:
  - Paclitaxel ± Trastuzumab  
  - Veliparib + Carboplatin  
  - AMG-386 (Trebananib) ± Trastuzumab  
  - Neratinib, Ganetespib, Ganitumab, Pembrolizumab, Pertuzumab, T-DM1  

#### Additional Information

- **Batch Correction**:  
  - **ComBat** was applied to correct batch effects in gene expression data.  
  - Adjusted ~800 patients; linear adjustments applied to remaining cases.  

- **Radiomics Data**:  
  - MRI dataset includes **4 serial DCE-MRI scans per patient**.  
  - Performed **before and during neoadjuvant chemotherapy** to track tumor volume changes and response.  

- **Data Availability**:  
  - Genomics: Available in [GEO: GSE194040](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE194040).  
  - Imaging: MRI datasets accessible via [The Cancer Imaging Archive (ISPY2 Collection)](https://www.cancerimagingarchive.net/collection/ispy2/).  

- **Key Reference**:  
  - Wolf DM et al., *Cancer Cell*, 2022. [PMID: 35623341](https://pubmed.ncbi.nlm.nih.gov/35623341).  

---

## Section 1: Data Processing 

### Load Libraries
 
```{r libraries, include=TRUE, message=FALSE, warning=FALSE}
library(here)
library(MultiAssayExperiment) 
library(knitr)
library(readxl)
library(ggplot2)
library(dplyr)
library(tidyr)
library(ggvenn)
library(kableExtra) 
library(limma)
library(EnhancedVolcano)
library(msigdbr)
library(fgsea)
library(patchwork)
library(stringr)
library(pheatmap)
library(stringr)
```

### Load Genomics Data {.tabset .tabset-pills}

Load MultiAssayExperiment `.rds` File.
This step extracts **clinical, expression (Microarray quantile normalized), and annotation data** for further analysis.

```{r load multiassay and extract Data}
# Define base directory
input_dir <- file.path(here(), "data/rawdata/genomics")
output_dir <- file.path(here(), "data/results/genomics")
dir.create(file.path(output_dir, "DEG_analysis"), recursive = TRUE, showWarnings = FALSE)
dir.create(file.path(output_dir, "pathway_analysis"), recursive = TRUE, showWarnings = FALSE)
dir.create(file.path(here(), "data/results/radiogenomics"), recursive = TRUE, showWarnings = FALSE)

# Load dataset
mae <- readRDS(file.path(input_dir, "ICB_Wolf.rds"))

# Extract clinical, expression, and annotation data
clin <- data.frame(colData(mae))  
expr <- assays(mae)[["expr"]] # dim 18348 x 987
annot <- data.frame(rowData(mae@ExperimentList$expr))

# Create a named vector to map gene_id to gene_name
gene_map <- setNames(annot$gene_name, annot$gene_id)
rownames(expr) <- gene_map[rownames(expr)]
rownames(annot) <- annot$gene_name
```

#### Clinical Data 

Display first 50 rows of Clinical Data.

```{r clin summery}
kable(head(clin, 50), caption = "Genomics Clinical Data") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE) %>%
  scroll_box(height = "400px")
```

#### Microarray Expression Data 

Display first 20 rows of quantile-normalized Microarray data.

```{r expr summery}
kable(head(expr, 20), caption = "Expression Data - Microarray") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE) %>%
  scroll_box(height = "400px")
```

#### Annotation Data 

Displays gene annotation for the dataset.

```{r data summery}
kable(head(annot, 20), caption = "Gene Annotation Data") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE) %>%
  scroll_box(height = "400px")
```

### Load Radiomics Data {.tabset .tabset-pills}

The radiomics dataset includes dynamic contrast-enhanced MRI (**DCE-MRI**) scans used to monitor tumor response to treatment.

**Metadata Source:**

* **File:** `ISPY2-Imaging-Cohort-1-Clinical-Data.xlsx`
* **Access:** [The Cancer Imaging Archive – I-SPY2 Collection](https://www.cancerimagingarchive.net/collection/ispy2/)
* **Total Patients:** **985** (719 I-SPY2 cases, 266 [ACRIN-6698 cases](https://www.cancerimagingarchive.net/collection/ispy1/))

**Radiomics (MRI) Data:**

* Each patient has **4 MRI scans** (T0–T3)
* MRI features include: Functional Tumor Volume (**FTV**), Longest Tumor Diameter, Sphericity
* Data collected from **22+ clinical centers**

```{r Load Radiomics Metadata}
Radiomics_clin <- read_excel(file.path(input_dir, "ISPY2-Imaging-Cohort-1-Clinical-Data.xlsx"))  

kable(Radiomics_clin, caption = "Radiomics Clinical Data") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE) %>%
  scroll_box(height = "400px")
```

### Match Patients: Genomics and Radiomics

Identify overlapping patients between genomics (gene expression) and radiomics (MRI scans) datasets.

```{r check Patient Matching}
# Prepare genomic clinical data
Genomics_clin <- clin  
Genomics_clin$patientid <- sub("^X", "", Genomics_clin$patientid)
rownames(Genomics_clin) <- Genomics_clin$patientid
colnames(expr) <- sub("^X", "", colnames(expr))

# Identify common patients
common_patients <- intersect(Genomics_clin$patientid, Radiomics_clin$Patient_ID)  
cat("Patients with both microarray and MRI data:", length(common_patients), "\n")

# Venn diagram
venn_data <- list(Genomics = Genomics_clin$patientid, Radiomics = Radiomics_clin$Patient_ID)

ggvenn(venn_data, fill_color = c("#A6CEE3", "#FB9A99"), stroke_size = 0.5, set_name_size = 4) +
  ggtitle("Overlap of Genomics (Microarray) and Radiomics (MRI) Patient IDs") +
  theme(plot.title = element_text(hjust = 0.5, size = 12, face = "bold"),
        plot.background = element_rect(fill = "white"),
        panel.background = element_blank())
```

### Merge Genomics & Radiomics Data

Merging both datasets based on **common patient IDs**.

```{r Merge}
# Add prefix to radiomics columns
colnames(Radiomics_clin)[-which(colnames(Radiomics_clin) == "Patient_ID")] <- 
  paste0("Radiomics_", colnames(Radiomics_clin)[-which(colnames(Radiomics_clin) == "Patient_ID")])

# Merge datasets
merged_clin <- merge(Genomics_clin, Radiomics_clin, by.x = "patientid", by.y = "Patient_ID")
merged_clin <- merged_clin[order(merged_clin$patientid), ]

# Save dataset
write.csv(merged_clin, 
          file = file.path(here(), "data/results/radiogenomics", "combined_clin_radiogenomic.csv"), 
          row.names = FALSE)

kable(head(merged_clin, 50), caption = "Merged Clinical Data (Genomics & Radiomics)") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE) %>%
  scroll_box(height = "400px")
```

### Segmentation Overview

We evaluated segmentation consistency by verifying that each patient had four **DCE-MRI** segmentation masks — one per time point — for reliable **Functional Tumor Volume (FTV)** analysis. Most patients met this requirement; those with fewer were flagged for review.

#### DCE-MRI Segmentation

Each scan undergoes multiple segmentation steps:  

1. **Manual VOI (Volume of Interest):** A rectangular region around the enhancing tumor.  
2. **Background Masking:** Removes noise and fat saturation artifacts.  
3. **Percent Enhancement (PE) Threshold:** Filters non-enhancing tissue.  
4. **Connectivity Filter:** Ensures 3D tumor voxel connectivity.  
5. **OMIT Regions:** Manual exclusion of non-tumor areas.  

Segmentation masks are stored as **2D DICOM images**, aligned with the original **DCE-MRI scans**.

> Download masks: [UCSF Box](https://ucsf.box.com/s/d09xetaatzgnz2p9bxlqh50rrzilgcoj)

#### Segmentation Metadata {.tabset .tabset-pills}

##### ISPY1: MRI Metadata  

Stored in `ispy1_dce_series_info.xlsx`, this file contains metadata for **221 patients**.

```{r ISPY1 segmentation metadata}
ispy1_dce_series_info <- read_excel(
  path = file.path(input_dir, "ispy1_dce_series_info.xlsx"), 
  sheet = "DCE Series, All 20160823"
) 

ispy1_dce_series_info <- ispy1_dce_series_info %>%
  mutate(TIME_POINT = recode(TIME_POINT, "T1" = "T0","T2" = "T1","T3" = "T2","T4" = "T3"))

kable(head(ispy1_dce_series_info, 20), caption = "DCE MRI Series Information") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE) %>%
  scroll_box(height = "400px")
```

##### ISPY2: MRI Metadata  

Metadata for ISPY2 was extracted from `series-data1741883133411.csv`, containing records for **719 patients**.

```{r segmentation metadata}
ispy2_dce_series_info <- read.csv(file.path(input_dir, "series-data1741883133411.csv"))
colnames(ispy2_dce_series_info)[colnames(ispy2_dce_series_info) == "Subject.ID"] <- "PATIENT_ID"
ispy2_dce_series_info$TIME_POINT <- gsub(".*(T[0-9]+).*", "\\1", ispy2_dce_series_info$Study.Description)

kable(head(ispy2_dce_series_info, 20), caption = "ISPY2 DCE MRI Series Information") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE) %>%
  scroll_box(height = "400px")

# Merge ISPY1 and ISPY2 metadata
# ispy1_selected <- ispy1_dce_series_info[, c("PATIENT_ID", "TIME_POINT")]
# ispy2_selected <- ispy2_dce_series_info[, c("PATIENT_ID", "TIME_POINT")]
ispy1and2 <- merge(ispy1_dce_series_info, ispy2_dce_series_info, by = c("PATIENT_ID", "TIME_POINT"), all = TRUE)

```

#### Segmentation Completeness Check

Each of the 940 patients is expected to have at least four segmentation masks, one for each of their four MRI scans.

```{r Check Segmentation}
series_per_patient <- ispy1and2 %>%
  group_by(PATIENT_ID, TIME_POINT) %>%
  summarize(Series_Count = n(), .groups = 'drop') %>%
  pivot_wider(names_from = TIME_POINT, values_from = Series_Count, values_fill = 0) %>%
  mutate(across(starts_with("T"), as.numeric),
         Total_Segmentations = rowSums(select(., starts_with("T")), na.rm = TRUE))

kable(series_per_patient, caption = "Segmentation Counts per Patient") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE) %>%
  scroll_box(height = "400px")

# Summary per timepoint
timepoints <- intersect(c("T0", "T1", "T2", "T3"), colnames(series_per_patient))
summary_df <- data.frame(
  Timepoint = timepoints,
  Present = sapply(timepoints, function(t) sum(series_per_patient[[t]] > 0)),
  Missing = sapply(timepoints, function(t) sum(series_per_patient[[t]] == 0))
)

kable(summary_df, caption = "Segmentation Availability per Timepoint") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE)


# Patients with both T0 and T1 (needed for delta-radiomics)
if (all(c("T0", "T1") %in% colnames(series_per_patient))) {
  has_both <- sum(series_per_patient$T0 > 0 & series_per_patient$T1 > 0)
  has_either <- sum(series_per_patient$T0 > 0 | series_per_patient$T1 > 0)
  cat("Patients with both T0 and T1:", has_both, "\n")
  cat("Patients with at least T0 or T1:", has_either, "\n")
}
```

Patients missing **one or more segmentation masks** need further review.

```{r missing segmentations}
patients_less_than_4 <- series_per_patient %>%
  filter(Total_Segmentations < 4) %>%
  select(PATIENT_ID, Total_Segmentations)

kable(patients_less_than_4, caption = "Patients with Less Than 4 Segmentation Masks") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE) %>%
  scroll_box(height = "400px")
```

---

## Section 2: Genomic Analysis

**Objective:** Process microarray expression data to identify predictive and biologically meaningful features associated with treatment response.

### 1. Subset Radiogenomics

982 patients shared between genomic and radiomics data; 18,348 genes.

```{r subset to Radiogenomics}
clin_data <- merged_clin
expr_data <- expr[, common_patients]
annot_data <- annot[annot$gene_name %in% common_patients, ]
```

### 2. Keep Protein-Coding Genes

* Retained only **protein-coding genes** (17,547) based on **GENCODE annotations**
* *Optional:* Apply variance filtering to remove low-variance genes -Skipped.

```{r restrict to protein-coding genes}
# Subset to common patients
expr_data <- expr[, common_patients]
annot_data <- annot[annot$gene_name %in% common_patients, ]
clin_data <- merged_clin[merged_clin$patientid %in% common_patients, ]

# Restrict to protein-coding genes
annot_proteincoding <- annot[annot$gene_type == "protein_coding", ]
expr_data <- expr_data[rownames(expr_data) %in% rownames(annot_proteincoding), ]

# Treatment frequency tables
DT::datatable(as.data.frame(table(merged_clin$treatment)), 
              colnames = c("Treatment", "Count"))
DT::datatable(as.data.frame(table(merged_clin$treatmentid)), 
              colnames = c("Treatment ID", "Count"))

# Expression density plot (all treatments)
plot(
  density(as.numeric(as.matrix(expr_data)), na.rm = TRUE),
  main = "Overall Expression Density (All treatments)",
  xlab = "Expression", ylab = "Density"
)

# Expression density by treatment group
treatments <- unique(na.omit(merged_clin$treatment))
for (treat in treatments) {
  patients <- merged_clin$patientid[merged_clin$treatment == treat]
  expr_subset <- expr_data[, colnames(expr_data) %in% patients, drop = FALSE]
  plot(
    density(as.numeric(as.matrix(expr_subset)), na.rm = TRUE),
    main = paste("Expression Density:", treat),
    xlab = "Expression", ylab = "Density"
  )
}
```

### 3. PredictIO Signatures

* Loaded **55 curated immune-oncology (IO) RNA signatures** (`.rda` format)
* Each signature derived from **peer-reviewed publications**
* Signatures used to assess **immune response**, **sensitivity**, or **resistance** to immune checkpoint blockade
* Mapped to **GENCODE v40**
* Sources: [bhklab/SignatureSets](https://github.com/bhklab/SignatureSets) | [predictio.ca](https://predictio.ca/)

```{r PredictIO sigs intersection}
sig_path <- file.path(input_dir, "predictIO_sigs")
rda_files <- list.files(sig_path, pattern = "\\.rda$", full.names = TRUE)
summary_list <- list()

for (file in rda_files) {
  sig_obj <- get(load(file))
  sig_name <- if ("signature_name" %in% colnames(sig_obj) && length(unique(sig_obj$signature_name)) > 0)
    unique(sig_obj$signature_name)[1]
  else
    tools::file_path_sans_ext(basename(file))

  if (!"gene_name" %in% colnames(sig_obj)) next

  genes <- sig_obj$gene_name
  matched <- intersect(rownames(expr_data), genes)

  summary_list[[sig_name]] <- data.frame(
    signature = sig_name,
    n_genes_total = length(genes),
    n_genes_matched = length(matched),
    matched_genes = paste(sort(matched), collapse = ", ")
  )
}

summary_df <- do.call(rbind, summary_list)

kable(summary_df, caption = "Summary of gene overlap between IO signatures and expression data") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE) %>%
  scroll_box(height = "400px")
```

### 4. DE Analysis (R vs NR) — Full Cohort

performs differential expression (DE) analysis using limma to find genes that differ between Responders vs Non-Responders across all 982 patients.

```{r DEG-Full dataset analysis, fig.width=7, fig.height=7}
# Design and model
design_full <- model.matrix(~ clin_data$response)
fit_full <- eBayes(lmFit(expr_data, design_full))
top_full <- topTable(fit_full, sort.by = "P", n = Inf)
top_full$Gene <- rownames(top_full)
fc_cutoff <- 1.5
p_cutoff <- 0.05
top_n_labels <- Inf

# Save results
write.table(top_full, file.path(output_dir, "DEG_analysis", "full_limma_fit.txt"),
            sep = "\t", quote = FALSE, row.names = FALSE)

DT::datatable(round(top_full[sapply(top_full, is.numeric)], 3), 
              caption = "TopTable - Full Cohort")

# Clean numeric columns for volcano
top_full.clean <- top_full[, sapply(top_full, is.numeric)]
rownames(top_full.clean) <- rownames(top_full)

# Labels: top 20 red dots only
red_genes_pval <- top_full.clean[
  top_full.clean$P.Value < p_cutoff & abs(top_full.clean$logFC) >= fc_cutoff,
]
red_genes_pval <- head(red_genes_pval[order(red_genes_pval$P.Value), ], top_n_labels)
selectLab_pval <- rownames(red_genes_pval)

red_genes_fdr <- top_full.clean[
  top_full.clean$adj.P.Val < p_cutoff & abs(top_full.clean$logFC) >= fc_cutoff,
]
red_genes_fdr <- head(red_genes_fdr[order(red_genes_fdr$adj.P.Val), ], top_n_labels)
selectLab_fdr <- rownames(red_genes_fdr)

# Counts
n_pval_full <- sum(top_full.clean$P.Value < p_cutoff & abs(top_full.clean$logFC) >= fc_cutoff)
n_fdr_full <- sum(top_full.clean$adj.P.Val < p_cutoff & abs(top_full.clean$logFC) >= fc_cutoff)

# Volcano plot: P-value
print(EnhancedVolcano(top_full.clean,
  lab = rownames(top_full.clean),
  selectLab = selectLab_pval,
  x = 'logFC', y = 'P.Value',
  pCutoff = p_cutoff, FCcutoff = fc_cutoff,
  xlim = c(-3, 3), ylim = c(0, 20),
  labSize = 3.5,
  labCol = "black",
  drawConnectors = TRUE,
  widthConnectors = 0.5,
  colConnectors = "grey50",
  maxoverlapsConnectors = Inf,
  title = "Volcano Plot (P value) - Full Cohort",
  subtitle = paste0("p < ", p_cutoff, " & |logFC| >= ", fc_cutoff, ": ", n_pval_full, " genes"),
  caption = paste0("Total genes: ", nrow(top_full.clean))
))

# Volcano plot: FDR
print(EnhancedVolcano(top_full.clean,
  lab = rownames(top_full.clean),
  selectLab = selectLab_fdr,
  x = 'logFC', y = 'adj.P.Val',
  pCutoff = p_cutoff, FCcutoff = fc_cutoff,
  xlim = c(-3, 3), ylim = c(0, 20),
  labSize = 3.5,
  labCol = "black",
  drawConnectors = TRUE,
  widthConnectors = 0.5,
  colConnectors = "grey50",
  maxoverlapsConnectors = Inf,
  legendLabels = c('Not sig.', 'Log FC', 'adj p-value', 'adj p-value & Log FC'),
  title = "Volcano Plot (FDR) - Full Cohort",
  subtitle = paste0("FDR < ", p_cutoff, " & |logFC| >= ", fc_cutoff, ": ", n_fdr_full, " genes"),
  caption = paste0("Total genes: ", nrow(top_full.clean))
))

# Summary
sig_genes_full <- rownames(top_full)[top_full$adj.P.Val < p_cutoff & abs(top_full$logFC) >= fc_cutoff]
cat("DE genes (adj.P.Val <", p_cutoff, "):", sum(top_full$adj.P.Val < p_cutoff), "\n")
cat("DE genes (adj.P.Val <", p_cutoff, "& |logFC| >=", fc_cutoff, "):", length(sig_genes_full), "\n")
cat("Significant gene names:\n")
print(sig_genes_full)
```

```{r}
# Summary table for full cohort
genes_full_fc1.0 <- rownames(top_full)[top_full$adj.P.Val < p_cutoff & abs(top_full$logFC) >= 1.0]
genes_full_fc1.5 <- rownames(top_full)[top_full$adj.P.Val < p_cutoff & abs(top_full$logFC) >= fc_cutoff]

summary_full <- data.frame(
  Cohort = "Full (982 patients)",
  FDR_sig = sum(top_full$adj.P.Val < p_cutoff),
  FDR_FC1.0 = length(genes_full_fc1.0),
  FDR_FC1.5 = length(genes_full_fc1.5),
  Top_Genes_FC1.0 = ifelse(length(genes_full_fc1.0) == 0, "–", paste(genes_full_fc1.0, collapse = ", ")),
  Top_Genes_FC1.5 = ifelse(length(genes_full_fc1.5) == 0, "–", paste(genes_full_fc1.5, collapse = ", "))
)

rownames(summary_full) <- NULL
colnames(summary_full) <- c("Cohort", "FDR < 0.05", 
                              "FDR & abs(logFC) >= 1.0", 
                              "FDR & abs(logFC) >= 1.5", 
                              "Genes (logFC >= 1.0)",
                              "Genes (logFC >= 1.5)")

kable(summary_full, row.names = FALSE,
      caption = "Summary of DE Genes - Full Cohort") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE) %>%
  scroll_box(height = "400px")

write.csv(summary_full, 
          file = file.path(output_dir, "DEG_analysis", "DE_summary_full_cohort.csv"),
          row.names = FALSE)
```

### 5. Pathway Enrichment — Full Cohort {.tabset .tabset-pills}

#### MSigDB Hallmark Gene Sets

* Gene sets from [MSigDB](https://www.gsea-msigdb.org/gsea/msigdb/human/collections.jsp#H)
* Analysis: **GSEA using the `fgsea` R package**
* Input: ranked gene list from differential expression results

```{r pathway analysis, fig.width=10, fig.height=7}
# Load hallmark gene sets
pathwaysDF <- msigdbr("Homo sapiens", category = "H")
pathwaysH <- split(as.character(pathwaysDF$gene_symbol), pathwaysDF$gs_name)

# Create ranked gene list
ranks <- top_full$logFC
names(ranks) <- top_full$Gene
ranks <- sort(ranks, decreasing = TRUE)

# Run GSEA
fgseaRes <- fgsea(pathways = pathwaysH, stats = ranks, nproc = 1)
fgseaRes <- na.omit(fgseaRes)

fgseaResTidy <- fgseaRes %>%
  as_tibble() %>%
  arrange(desc(NES)) %>%
  mutate(leadingEdge = sapply(leadingEdge, paste, collapse = ";"))
fgseaResTidy <- fgseaResTidy[order(fgseaResTidy$padj), ]

kable(fgseaResTidy) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE) %>%
  scroll_box(height = "400px")

write.table(fgseaResTidy, file.path(output_dir, "pathway_analysis", "full_fgseaRes_HRank.txt"),
            sep = "\t", row.names = FALSE, quote = FALSE)

# Significance colors
fgseaResTidy$adjPvalue <- ifelse(fgseaResTidy$padj <= 0.05, "significant", "non-significant")
cols <- c("non-significant" = "grey", "significant" = "red")

# NES barplot
ggplot(fgseaResTidy, aes(reorder(pathway, NES), NES, fill = adjPvalue)) +
  geom_col() +
  scale_fill_manual(values = cols) +
  coord_flip() +
  labs(x = "Pathway", y = "Normalized Enrichment Score",
       title = "Hallmark Pathways NES - Full Cohort")

# GSEA table plot
plotGseaTable(pathwaysH[fgseaRes$pathway[fgseaRes$padj < 0.05]], ranks, fgseaRes, gseaParam = 0.5)

# Dot plot
ggplot(fgseaResTidy[fgseaResTidy$padj < 0.05, ], aes(y = reorder(pathway, size), x = NES)) +
  geom_point(aes(color = padj, size = size)) +
  scale_color_gradientn(colours = c("red", "blue")) +
  labs(x = 'Normalized Enrichment Score', y = NULL) +
  theme(axis.title = element_text(face = 'bold'), axis.text = element_text(face = 'bold'))
```

#### KEGG Pathway Analysis {.tabset .tabset-pills}

##### KEGG_MEDICUS

```{r pathway analysis dotplot, fig.width=16, fig.height=10}
pathwaysKEGG_MED <- gmtPathways(file.path(input_dir, "c2.cp.kegg_medicus.v2023.2.Hs.symbols.gmt"))

fgseaRes_kegg <- fgsea(pathways = pathwaysKEGG_MED, stats = ranks, nproc = 1)

fgseaResTidy_kegg <- fgseaRes_kegg %>%
  as_tibble() %>%
  arrange(desc(NES)) %>%
  mutate(leadingEdge = sapply(leadingEdge, paste, collapse = ";"))

kable(fgseaResTidy_kegg) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE) %>%
  scroll_box(height = "400px")

write.table(fgseaResTidy_kegg, file.path(output_dir, "pathway_analysis", "full_fgseaRes_kegRank.txt"),
            sep = "\t", row.names = FALSE, quote = FALSE)

fgseaResTidy_kegg$adjPvalue <- ifelse(fgseaResTidy_kegg$padj <= 0.05, "significant", "non-significant")

ggplot(fgseaResTidy_kegg[fgseaResTidy_kegg$padj < 0.05, ], aes(reorder(pathway, NES), NES, fill = adjPvalue)) +
  geom_col() +
  scale_fill_manual(values = cols) +
  coord_flip() +
  labs(x = "Pathway", y = "Normalized Enrichment Score",
       title = "KEGG Pathways NES - Full Cohort")

ggplot(fgseaResTidy_kegg[fgseaResTidy_kegg$padj < 0.05, ], aes(y = reorder(pathway, size), x = NES)) +
  geom_point(aes(color = padj, size = size)) +
  scale_color_gradientn(colours = c("red", "blue")) +
  labs(x = 'Normalized Enrichment Score', y = NULL) +
  theme(axis.title = element_text(face = 'bold'), axis.text = element_text(face = 'bold'))
```

#### Over-Representation Analysis (ORA)

Identifies whether certain gene sets are over-represented in a target gene group.

```{r ORA full toptable, fig.width=18, fig.height=12}
deGenes <- rownames(top_full)[top_full$pval < 0.05]
geneUniverse <- rownames(expr_data)

# ORA for Hallmark pathways
oraHallmark <- fora(pathways = pathwaysH, genes = deGenes, universe = geneUniverse, minSize = 15, maxSize = 500)
oraHallmark$Pathway_Set <- "Hallmark"
oraHallmark$Significance <- ifelse(oraHallmark$pval < 0.05, "Significant", "Not Significant")

# ORA for KEGG_MEDICUS pathways
oraKeggMed <- fora(pathways = pathwaysKEGG_MED, genes = deGenes, universe = geneUniverse, minSize = 15, maxSize = 500)
oraKeggMed$Pathway_Set <- "KEGG_MEDICUS"
oraKeggMed$Significance <- ifelse(oraKeggMed$pval < 0.05, "Significant", "Not Significant")

kable(oraHallmark, caption = "Hallmark Pathway Enrichment Results") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE) %>%
  scroll_box(height = "400px")

kable(oraKeggMed, caption = "KEGG Medicus Pathway Enrichment Results") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE) %>%
  scroll_box(height = "400px")

write.table(transform(oraHallmark, overlapGenes = sapply(overlapGenes, paste, collapse = ",")),
            file = file.path(output_dir, "pathway_analysis", "Hallmark_ORA_Results.csv"),
            sep = "\t", row.names = FALSE, quote = FALSE)

write.table(transform(oraKeggMed, overlapGenes = sapply(overlapGenes, paste, collapse = ",")),
            file = file.path(output_dir, "pathway_analysis", "KeggMed_ORA_Results.csv"),
            sep = "\t", row.names = FALSE, quote = FALSE)
```

### 6. DE by Treatment (R vs NR) 

Treatment groups: `"chemo"`, `"chemo+targeted"`, `"IO+chemo"`, `"targeted"`

```{r DEG-subset dataset analysis,fig.width=7, fig.height=7}
DT::datatable(as.data.frame(table(merged_clin$treatment)), colnames = c("Treatment", "Count"))
DT::datatable(as.data.frame(table(merged_clin$treatmentid)), colnames = c("Treatment ID", "Count"))

treatment_list <- c("chemo", "chemo+targeted", "IO+chemo", "targeted")
sig_gene_list <- list()
top_table_list <- list()

# Define thresholds once — change here to update everywhere
fc_cutoff <- 1.5
p_cutoff <- 0.05
top_n_labels <- Inf 

for (treatment in treatment_list) {
  cat("\nProcessing:", treatment, "\n")
  
  # Subset data
  clin_sub <- clin_data[clin_data$treatment == treatment, ]
  matched_ids <- clin_sub$patientid[clin_sub$patientid %in% colnames(expr_data)]
  expr_sub <- expr_data[, matched_ids]

  # Design and fit
  design <- model.matrix(~ clin_sub$response)
  fit <- lmFit(expr_sub, design)
  fit <- eBayes(fit)
  top.table <- topTable(fit, sort.by = "P", n = Inf)

  cat("DE genes (adj.P.Val <", p_cutoff, "):", sum(top.table$adj.P.Val < p_cutoff), "\n")

  top.table.clean <- top.table[, sapply(top.table, is.numeric)]
  rownames(top.table.clean) <- rownames(top.table)

  # Labels: top N red dots only
  red_genes_pval <- top.table.clean[
    top.table.clean$P.Value < p_cutoff & abs(top.table.clean$logFC) >= fc_cutoff,
  ]
  red_genes_pval <- head(red_genes_pval[order(red_genes_pval$P.Value), ], top_n_labels)
  selectLab_pval <- rownames(red_genes_pval)

  red_genes_fdr <- top.table.clean[
    top.table.clean$adj.P.Val < p_cutoff & abs(top.table.clean$logFC) >= fc_cutoff,
  ]
  red_genes_fdr <- head(red_genes_fdr[order(red_genes_fdr$adj.P.Val), ], top_n_labels)
  selectLab_fdr <- rownames(red_genes_fdr)

  # Counts for subtitle
  n_pval <- sum(top.table.clean$P.Value < p_cutoff & abs(top.table.clean$logFC) >= fc_cutoff)
  n_fdr <- sum(top.table.clean$adj.P.Val < p_cutoff & abs(top.table.clean$logFC) >= fc_cutoff)

  # Volcano plot: P-value
  print(EnhancedVolcano(top.table.clean,
    lab = rownames(top.table.clean),
    selectLab = selectLab_pval,
    x = 'logFC', y = 'P.Value',
    pCutoff = p_cutoff, FCcutoff = fc_cutoff,
    xlim = c(-5, 5), ylim = c(0, 8),
    labSize = 3.5,
    labCol = "black",
    drawConnectors = TRUE,
    widthConnectors = 0.5,
    colConnectors = "grey50",
    maxoverlapsConnectors = Inf,
    title = paste("Volcano Plot (P value) -", treatment),
    subtitle = paste0("p < ", p_cutoff, " & |logFC| >= ", fc_cutoff, ": ", n_pval, " genes (top ", top_n_labels, " labeled)"),
    caption = paste0("Total genes: ", nrow(top.table.clean))
  ))

  # Volcano plot: FDR
  print(EnhancedVolcano(top.table.clean,
    lab = rownames(top.table.clean),
    selectLab = selectLab_fdr,
    x = 'logFC', y = 'adj.P.Val',
    pCutoff = p_cutoff, FCcutoff = fc_cutoff,
    xlim = c(-5, 5), ylim = c(0, 8),
    labSize = 3.5,
    labCol = "black",
    drawConnectors = TRUE,
    widthConnectors = 0.5,
    colConnectors = "grey50",
    maxoverlapsConnectors = Inf,
    legendLabels = c('Not sig.', 'Log FC', 'adj p-value', 'adj p-value & Log FC'),
    title = paste("Volcano Plot (FDR) -", treatment),
    subtitle = paste0("FDR < ", p_cutoff, " & |logFC| >= ", fc_cutoff, ": ", n_fdr, " genes (top ", top_n_labels, " labeled)"),
    caption = paste0("Total genes: ", nrow(top.table.clean))
  ))

  # Save and display
  top.table$Gene <- rownames(top.table)
  file_name <- paste0(gsub("\\+", "_", treatment), "_limma_fit.txt")
  write.table(top.table, file = file.path(output_dir, "DEG_analysis", file_name),
              row.names = FALSE, sep = "\t", quote = FALSE)

  top.display <- top.table
  top.display[, sapply(top.display, is.numeric)] <- round(top.display[, sapply(top.display, is.numeric)], 3)
  print(DT::datatable(top.display, caption = paste("TopTable -", treatment)))

  # Report
  sig_genes <- rownames(top.table)[top.table$adj.P.Val < p_cutoff & abs(top.table$logFC) >= fc_cutoff]
  cat("DE genes (FDR <", p_cutoff, "& |logFC| >=", fc_cutoff, "):", length(sig_genes), "\n")
  cat("Significant gene names:\n")
  print(sig_genes)

  top_table_list[[treatment]] <- top.table
  sig_gene_list[[treatment]] <- sig_genes

  gene_file <- paste0(gsub("\\+", "_", treatment), "_significant_genes.txt")
  writeLines(sig_genes, file.path(output_dir, "DEG_analysis", gene_file))
}
```


**Summary of Differentially Expressed Genes by Treatment**:

**IO+Chemo Key Findings (47 DE genes):**

- **Upregulated in responders (immune-hot):** 
  LAMP3, CCL13, IDO1, CXCL13, CXCL9, MARCO, UBD, CHI3L2, CRABP1, KRT5
  → Immune activation: dendritic cells, T-cell chemokines, interferon-induced genes

- **Upregulated in non-responders (hormone-driven):** 
  CA12, AGR2, FOXA1, AR, ACOX2, GATA3, TFF1, TFF3, SCUBE2, MSMB
  → Estrogen receptor pathway: luminal markers, hormone-regulated genes

- **Key takeaway:** Immune-hot tumors respond to pembrolizumab + chemo; 
  cold, hormone-driven tumors do not.

```{r define sub_treat}
# Summary table with both thresholds
summary_rows <- list()
for (treatment in treatment_list) {
  tt <- top_table_list[[treatment]]
  
  genes_fc1.0 <- rownames(tt)[tt$adj.P.Val < p_cutoff & abs(tt$logFC) >= 1.0]
  genes_fc1.5 <- rownames(tt)[tt$adj.P.Val < p_cutoff & abs(tt$logFC) >= fc_cutoff]
  
  summary_rows[[treatment]] <- data.frame(
    Treatment = treatment,
    FDR_sig = sum(tt$adj.P.Val < p_cutoff),
    FDR_FC1.0 = length(genes_fc1.0),
    FDR_FC1.5 = length(genes_fc1.5),
    Top_Genes_FC1.0 = ifelse(length(genes_fc1.0) == 0, "–", paste(genes_fc1.0, collapse = ", ")),
    Top_Genes_FC1.5 = ifelse(length(genes_fc1.5) == 0, "–", paste(genes_fc1.5, collapse = ", "))
  )
}

summary_df <- do.call(rbind, summary_rows)
rownames(summary_df) <- NULL
colnames(summary_df) <- c("Treatment", "FDR < 0.05", 
                           "FDR & abs(logFC) >= 1.0", 
                           "FDR & abs(logFC) >= 1.5", 
                           "Genes (logFC >= 1.0)",
                           "Genes (logFC >= 1.5)")

kable(summary_df, row.names = FALSE, 
      caption = "Summary of DE Genes by Treatment") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE) %>%
  scroll_box(height = "400px")

write.csv(summary_df, 
          file = file.path(output_dir, "DEG_analysis", "DE_summary_by_treatment.csv"),
          row.names = FALSE)
```



### 7. Enrichment by Treatment {.tabset .tabset-pills}

> **Note:** Pathway enrichment is focused on IO+Chemo and Targeted arms, as Chemo (0 DE genes) and Chemo+Targeted (1 DE gene) showed minimal differential expression signal.

```{r define sub_treats}
sub_treat <- c("IO+chemo", "targeted", "chemo+targeted")
```

#### MSigDB Hallmark Gene Sets

```{r per-treat hallmark, fig.width=7, fig.height=7}
fgsea_result_list <- list()
fgsea_plot_list <- list()
fgsea_dotplot_list <- list()

for (treatment in sub_treat) {
  message("Running GSEA for: ", treatment)
  
  top_treat <- top_table_list[[treatment]]
  ranks <- top_treat$logFC
  names(ranks) <- top_treat$Gene
  ranks <- sort(ranks, decreasing = TRUE)

  fgseaRes <- fgsea(pathways = pathwaysH, stats = ranks, nproc = 1)
  fgseaRes <- na.omit(fgseaRes)
  
  fgseaResTidy <- fgseaRes %>%
    as_tibble() %>%
    arrange(desc(NES)) %>%
    mutate(
      leadingEdge = sapply(leadingEdge, paste, collapse = ";"),
      adjPvalue = ifelse(padj <= 0.05, "significant", "non-significant")
    )

  fgsea_result_list[[treatment]] <- fgseaResTidy

  # Barplot (all pathways)
  p_bar <- ggplot(fgseaResTidy, aes(reorder(pathway, NES), NES, fill = adjPvalue)) +
    geom_col() +
    scale_fill_manual(values = c("significant" = "red", "non-significant" = "grey")) +
    coord_flip() +
    theme_minimal() +
    labs(title = paste("Hallmark NES -", treatment), x = "Pathway", y = "Normalized Enrichment Score")

  fgsea_plot_list[[treatment]] <- p_bar
  print(p_bar)

  # Dot plot (significant only)
  sig_fgsea <- fgseaResTidy %>% filter(padj < p_cutoff)

  if (nrow(sig_fgsea) > 0) {
    p_dot <- ggplot(sig_fgsea, aes(x = NES, y = reorder(pathway, NES))) +
      geom_point(aes(color = padj, size = size)) +
      scale_color_gradientn(colours = c("red", "blue")) +
      theme_minimal() +
      labs(x = "Normalized Enrichment Score", y = NULL,
           title = paste("Significant Hallmark Pathways -", treatment),
           subtitle = paste0("FDR < ", p_cutoff, ": ", nrow(sig_fgsea), " pathways")) +
      theme(axis.title = element_text(face = 'bold'),
            axis.text = element_text(face = 'bold'))

    fgsea_dotplot_list[[treatment]] <- p_dot
    print(p_dot)
  } else {
    cat("No significant pathways for", treatment, "\n")
  }

  # Table
  print(
    kable(fgseaResTidy, caption = paste("fgsea results -", treatment)) %>%
      kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE) %>%
      scroll_box(height = "400px")
  )

  # Save results
  write.table(fgseaResTidy,
    file.path(output_dir, "pathway_analysis", paste0(gsub("\\+", "_", treatment), "_fgseaRes_Hallmark.txt")),
    sep = "\t", row.names = FALSE, quote = FALSE)
}
```

#### KEGG_MEDICUS Pathway Analysis

```{r per-treat Kegg dotplot, fig.width=13, fig.height=10}
fgsea_kegg_result_list <- list()
fgsea_kegg_barplot_list <- list()
fgsea_kegg_dotplot_list <- list()

for (treatment in sub_treat) {
  message("Running KEGG GSEA for: ", treatment)

  top_treat <- top_table_list[[treatment]]
  ranks <- setNames(top_treat$logFC, top_treat$Gene)
  ranks <- sort(ranks, decreasing = TRUE)

  fgseaRes_kegg <- fgsea(pathways = pathwaysKEGG_MED, stats = ranks, nproc = 1)
  fgseaRes_kegg <- na.omit(fgseaRes_kegg)

  fgseaResTidy_kegg <- fgseaRes_kegg %>%
    as_tibble() %>%
    arrange(desc(NES)) %>%
    mutate(
      leadingEdge = sapply(leadingEdge, paste, collapse = ";"),
      adjPvalue = ifelse(padj <= 0.05, "significant", "non-significant")
    )

  fgsea_kegg_result_list[[treatment]] <- fgseaResTidy_kegg

  write.table(fgseaResTidy_kegg,
    file.path(output_dir, "pathway_analysis", paste0(gsub("\\+", "_", treatment), "_fgseaRes_kegg.txt")),
    sep = "\t", row.names = FALSE, quote = FALSE)

  sig_fgsea <- fgseaResTidy_kegg %>% filter(padj <= p_cutoff)

  # Barplot (significant only)
  if (nrow(sig_fgsea) > 0) {
    p_bar <- ggplot(sig_fgsea, aes(reorder(pathway, NES), NES, fill = adjPvalue)) +
      geom_col() +
      scale_fill_manual(values = c("significant" = "red")) +
      coord_flip() +
      theme_minimal() +
      labs(x = "Pathway", y = "NES", 
           title = paste("KEGG NES -", treatment),
           subtitle = paste0("FDR < ", p_cutoff, ": ", nrow(sig_fgsea), " pathways"))

    fgsea_kegg_barplot_list[[treatment]] <- p_bar
    print(p_bar)

    # Dot plot (significant only)
    p_dot <- ggplot(sig_fgsea, aes(x = NES, y = reorder(pathway, NES))) +
      geom_point(aes(color = padj, size = size)) +
      scale_color_gradientn(colours = c("red", "blue")) +
      theme_minimal() +
      labs(x = "Normalized Enrichment Score", y = NULL, 
           title = paste("KEGG Dot Plot -", treatment),
           subtitle = paste0("FDR < ", p_cutoff, ": ", nrow(sig_fgsea), " pathways")) +
      theme(axis.title = element_text(face = 'bold'), 
            axis.text = element_text(face = 'bold'))

    fgsea_kegg_dotplot_list[[treatment]] <- p_dot
    print(p_dot)
  } else {
    cat("No significant KEGG pathways for", treatment, "\n")
  }

  # Table
  print(
    kable(fgseaResTidy_kegg, caption = paste("KEGG GSEA results -", treatment)) %>%
      kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE) %>%
      scroll_box(height = "400px")
  )
}
```

#### Over-Representation Analysis (per-treatment)

```{r per-treat ORA, fig.width=18, fig.height=12}
ora_hallmark_result_list <- list()
ora_keggmed_result_list <- list()

for (treatment in sub_treat) {
  message("Running ORA for: ", treatment)

  top_treat <- top_table_list[[treatment]]
  deGenes <- rownames(top_treat)[top_treat$pval < 0.05]
  geneUniverse <- rownames(expr_data)

  # ORA Hallmark
  oraHallmark <- fora(pathways = pathwaysH, genes = deGenes, universe = geneUniverse, minSize = 15, maxSize = 500)
  oraHallmark$Pathway_Set <- "Hallmark"
  oraHallmark$Significance <- ifelse(oraHallmark$pval < 0.05, "Significant", "Not Significant")
  oraHallmark$Treatment <- treatment
  ora_hallmark_result_list[[treatment]] <- oraHallmark

  # ORA KEGG_MEDICUS
  oraKeggMed <- fora(pathways = pathwaysKEGG_MED, genes = deGenes, universe = geneUniverse, minSize = 15, maxSize = 500)
  oraKeggMed$Pathway_Set <- "KEGG_MEDICUS"
  oraKeggMed$Significance <- ifelse(oraKeggMed$pval < 0.05, "Significant", "Not Significant")
  oraKeggMed$Treatment <- treatment
  ora_keggmed_result_list[[treatment]] <- oraKeggMed

  # Save
  write.table(transform(oraHallmark, overlapGenes = sapply(overlapGenes, paste, collapse = ",")),
    file = file.path(output_dir, "pathway_analysis", paste0(gsub("\\+", "_", treatment), "_Hallmark_ORA.tsv")),
    sep = "\t", row.names = FALSE, quote = FALSE)

  write.table(transform(oraKeggMed, overlapGenes = sapply(overlapGenes, paste, collapse = ",")),
    file = file.path(output_dir, "pathway_analysis", paste0(gsub("\\+", "_", treatment), "_KeggMed_ORA.tsv")),
    sep = "\t", row.names = FALSE, quote = FALSE)

  print(kable(head(oraHallmark, 10), caption = paste("Top Hallmark ORA -", treatment)) %>%
    kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE))
}
```

### 8. Gene–Response Correlation

Does the expression of a specific gene predict whether a patient will respond to treatment or not? Sections 4–7 looked at genes and pathways as groups. This section zooms in and tests **each gene individually** to see if higher expression is associated with response.

### 8. Gene–Response Correlation

Does the expression of a specific gene predict whether a patient will respond to treatment or not? Sections 4–7 looked at genes and pathways as groups. This section zooms in and tests **each gene individually** to see if higher expression is associated with response, then groups them into biologically meaningful combinations.

**Candidate genes from two sources:**

**1. DE Significant Genes (from Section 6):**
* **IO+Chemo:** 47 genes at |logFC| ≥ 1.5 (143 at |logFC| ≥ 1.0)
* **Targeted:** 3 genes at |logFC| ≥ 1.5 (5 at |logFC| ≥ 1.0)
* Gene-response correlation uses the broader |logFC| ≥ 1.0 list to capture moderate but real effects

**2. Key Biological Candidates (from literature):**
* **Immune:** CD8A, CD8B, PDCD1 (PD-1), CD274 (PD-L1), IFNG, GZMB, PRF1, CXCL9
* **Luminal/Resistance:** ESR1, PGR, ERBB2

> **Note:** CD8A/CD8B were selected based on [Sun et al. (PMID: 30120041)](https://pubmed.ncbi.nlm.nih.gov/30120041/), which used CD8B to build a radiomic-immune signature linking CT features to CD8 cell infiltration. PDCD1/CD274 are the direct targets of the IO drugs (pembrolizumab) used in I-SPY2. ESR1/PGR/ERBB2 connect to the estrogen response and HER2 pathways identified in Sections 5–7.

**PredictIO Signatures (from Section 3):**
* 55 curated immune-oncology RNA signatures — used as **validation only**
* Not included as candidate source (derived from other studies/cancers)
* Used to check: do known IO signatures overlap with our DE-derived candidates?

**Analysis flow:**
1. Build candidate gene list (DE at |logFC| ≥ 1.0 + literature)
2. Test each gene individually against response (Wilcoxon, per treatment)
3. Group genes into biological combinations (immune, luminal, HER2) and test group scores
4. Validate against PredictIO signatures (overlap check)
5. Visualize: boxplots and heatmaps

```{r build candidate genes}
# DE genes at |logFC| >= 1.0 from IO+chemo and targeted 
de_genes_fc1 <- c()
for (treatment in c("IO+chemo", "targeted")) {
  tt <- top_table_list[[treatment]]
  genes <- rownames(tt)[tt$adj.P.Val < p_cutoff & abs(tt$logFC) >= 1.0]
  de_genes_fc1 <- c(de_genes_fc1, genes)
}
de_genes_fc1 <- unique(de_genes_fc1)

#  Literature genes 
literature_genes <- c("CD8A", "CD8B", "PDCD1", "CD274", "IFNG", "GZMB", "PRF1",
                       "CXCL9", "ESR1", "PGR", "ERBB2")

# Combine 
candidate_genes <- unique(c(de_genes_fc1, literature_genes))
candidate_genes <- candidate_genes[candidate_genes %in% rownames(expr_data)]

cat("Candidate genes:\n")
cat("  DE |logFC| >= 1.5 (IO+chemo + targeted):", 
    length(unique(c(sig_gene_list[["IO+chemo"]], sig_gene_list[["targeted"]]))), "\n")
cat("  DE |logFC| >= 1.0 (IO+chemo + targeted):", length(de_genes_fc1), "\n")
cat("  Literature:", length(literature_genes), "\n")
cat("  Total unique:", length(candidate_genes), "\n")

# Source tracking table 
source_df <- data.frame(
  Gene = candidate_genes,
  DE_FC1.5 = ifelse(candidate_genes %in% unique(c(sig_gene_list[["IO+chemo"]], sig_gene_list[["targeted"]])), "Yes", ""),
  DE_FC1.0 = ifelse(candidate_genes %in% de_genes_fc1, "Yes", ""),
  Literature = ifelse(candidate_genes %in% literature_genes, "Yes", ""),
  stringsAsFactors = FALSE
)
source_df <- source_df[order(source_df$DE_FC1.5, decreasing = TRUE), ]

kable(source_df, caption = "Candidate Gene Sources (|logFC| >= 1.0 for correlation)",
      row.names = FALSE) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE) %>%
  scroll_box(height = "400px")

write.csv(source_df,
          file = file.path(output_dir, "DEG_analysis", "candidate_gene_sources.csv"),
          row.names = FALSE)
```

#### Step 1: Single Gene–Response Correlation {.tabset .tabset-pills}

Test each candidate gene individually using Wilcoxon rank-sum test (R vs NR) per treatment arm.

```{r Gene correlation with response, warning=FALSE}
results <- list()
for (treat in sub_treat) {
  clin_sub <- clin_data[clin_data$treatment == treat, ]
  ids <- clin_sub$patientid[clin_sub$patientid %in% colnames(expr_data)]
  
  for (gene in candidate_genes) {
    expr_vals <- as.numeric(expr_data[gene, ids])
    response <- clin_sub$response[clin_sub$patientid %in% ids]
    
    wt <- wilcox.test(expr_vals ~ response)
    
    results[[paste(treat, gene)]] <- data.frame(
      Treatment = treat,
      Gene = gene,
      p_value = wt$p.value,
      mean_R = mean(expr_vals[response == "R"], na.rm = TRUE),
      mean_NR = mean(expr_vals[response == "NR"], na.rm = TRUE)
    )
  }
}

gene_response_df <- do.call(rbind, results)
gene_response_df$FDR <- p.adjust(gene_response_df$p_value, method = "BH")
gene_response_df$Direction <- ifelse(gene_response_df$mean_R > gene_response_df$mean_NR, "Up in R", "Up in NR")
gene_response_df <- gene_response_df[order(gene_response_df$p_value), ]

kable(gene_response_df, caption = "Gene-Response Correlation (Single Genes)") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE) %>%
  scroll_box(height = "400px")

write.csv(gene_response_df,
          file = file.path(output_dir, "DEG_analysis", "gene_response_correlation.csv"),
          row.names = FALSE)
```


```{r Boxplots single genes, fig.width=10, fig.height=8}
for (treat in sub_treat) {
  top_genes_df <- gene_response_df %>%
    filter(Treatment == treat, FDR < 0.05) %>%
    head(6)
  
  if (nrow(top_genes_df) == 0) next
  
  clin_sub <- clin_data[clin_data$treatment == treat, ]
  ids <- clin_sub$patientid[clin_sub$patientid %in% colnames(expr_data)]
  
  plot_list <- list()
  for (j in seq_len(nrow(top_genes_df))) {
    gene <- top_genes_df$Gene[j]
    fdr_val <- top_genes_df$FDR[j]
    
    df <- data.frame(
      Expression = as.numeric(expr_data[gene, ids]),
      Response = clin_sub$response[clin_sub$patientid %in% ids]
    )
    plot_list[[gene]] <- ggplot(df, aes(x = Response, y = Expression, fill = Response)) +
      geom_boxplot(outlier.shape = NA) +
      geom_jitter(width = 0.2, alpha = 0.5, size = 1) +
      scale_fill_manual(values = c("R" = "#E74C3C", "NR" = "#3498DB")) +
      labs(title = gene,
           subtitle = paste0("FDR = ", formatC(fdr_val, format = "e", digits = 2)),
           y = "Expression", x = NULL) +
      theme_minimal() +
      theme(legend.position = "none",
            plot.subtitle = element_text(size = 7))
  }
  
  combined <- wrap_plots(plot_list, ncol = 3) +
    plot_annotation(title = paste("Top Individual Genes -", treat))
  print(combined)
}
```

#### Step 2: Data-Driven Gene Group Scores {.tabset .tabset-pills}

Group candidate genes by biological function (from DE results in Section 6) and test combined z-score as a group signature.

```{r data-driven gene combinations, fig.width=10, fig.height=5}
# Define gene groups from our DE results 
immune_genes <- intersect(c("CD8A", "CD8B", "PDCD1", "CD274", "IFNG", "GZMB", "PRF1",
                             "CXCL9", "CXCL13", "IDO1", "LAMP3", "CCL13", "MARCO", "UBD",
                             "CHI3L2", "CRABP1", "KRT5"), candidate_genes)

luminal_genes <- intersect(c("ESR1", "PGR", "FOXA1", "GATA3", "AR", "AGR2", "CA12",
                              "TFF1", "TFF3", "SCUBE2", "ACOX2", "MSMB"), candidate_genes)

her2_genes <- intersect(c("ERBB2", "GRB7", "MNX1", "PNMT"), candidate_genes)

gene_groups <- list(
  Immune_Activation = immune_genes,
  Luminal_Resistance = luminal_genes,
  HER2_Associated = her2_genes
)
gene_groups <- gene_groups[sapply(gene_groups, length) >= 2]

cat("Gene groups:\n")
for (g in names(gene_groups)) {
  cat(" ", g, "(", length(gene_groups[[g]]), "genes):",
      paste(gene_groups[[g]], collapse = ", "), "\n")
}

# Test each group per treatment 
group_results <- list()
for (group_name in names(gene_groups)) {
  genes <- gene_groups[[group_name]]
  for (treat in sub_treat) {
    clin_sub <- clin_data[clin_data$treatment == treat, ]
    ids <- clin_sub$patientid[clin_sub$patientid %in% colnames(expr_data)]
    
    z_mat <- t(scale(t(expr_data[genes, ids])))
    group_score <- colMeans(z_mat, na.rm = TRUE)
    response <- clin_sub$response[clin_sub$patientid %in% ids]
    
    wt <- wilcox.test(group_score ~ response)
    
    group_results[[paste(treat, group_name)]] <- data.frame(
      Treatment = treat, Gene_Group = group_name,
      n_genes = length(genes),
      genes = paste(genes, collapse = ", "),
      p_value = wt$p.value,
      mean_R = mean(group_score[response == "R"], na.rm = TRUE),
      mean_NR = mean(group_score[response == "NR"], na.rm = TRUE)
    )
  }
}

group_score_df <- do.call(rbind, group_results)
group_score_df$FDR <- p.adjust(group_score_df$p_value, method = "BH")
group_score_df$Direction <- ifelse(group_score_df$mean_R > group_score_df$mean_NR, "Up in R", "Up in NR")
group_score_df <- group_score_df[order(group_score_df$p_value), ]

kable(group_score_df, caption = "Gene Group Scores vs Response", row.names = FALSE) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE) %>%
  scroll_box(height = "400px")

write.csv(group_score_df,
          file = file.path(output_dir, "DEG_analysis", "gene_group_scores_vs_response.csv"),
          row.names = FALSE)

wrap_genes <- function(genes, width = 55) {
  stringr::str_wrap(paste(genes, collapse = ", "), width = width)
}

# Boxplots for gene groups (with gene names in subtitle)
for (treat in sub_treat) {
  treat_groups <- group_score_df %>% filter(Treatment == treat)
  if (nrow(treat_groups) == 0) next

  clin_sub <- clin_data[clin_data$treatment == treat, ]
  ids <- clin_sub$patientid[clin_sub$patientid %in% colnames(expr_data)]
  if (length(ids) < 2) next

  plot_list <- list()
  for (i in seq_len(nrow(treat_groups))) {
    gname <- treat_groups$Gene_Group[i]
    fdr_val <- treat_groups$FDR[i]
    genes <- gene_groups[[gname]]
    if (length(genes) < 2) next

    z_mat <- t(scale(t(expr_data[genes, ids, drop = FALSE])))
    df <- data.frame(
      Score = colMeans(z_mat, na.rm = TRUE),
      Response = clin_sub$response[match(ids, clin_sub$patientid)]
    )

    genes_txt <- wrap_genes(genes, width = 55)

    plot_list[[gname]] <- ggplot(df, aes(x = Response, y = Score, fill = Response)) +
      geom_boxplot(outlier.shape = NA) +
      geom_jitter(width = 0.2, alpha = 0.5, size = 1) +
      scale_fill_manual(values = c("R" = "#E74C3C", "NR" = "#3498DB")) +
      labs(
        title = gname,
        subtitle = paste0(
          "FDR = ", formatC(fdr_val, format = "e", digits = 2),
          " | ", length(genes), " genes\n",
          genes_txt
        ),
        y = "Mean Z-Score",
        x = NULL
      ) +
      theme_minimal() +
      theme(
        legend.position = "none",
        plot.subtitle = element_text(size = 6, lineheight = 0.95),
        plot.title = element_text(size = 9)
      )
  }

  if (length(plot_list) == 0) next

  combined <- wrap_plots(plot_list, ncol = 3) +
    plot_annotation(title = paste("Gene Group Scores -", treat))

  print(combined)
}
```

#### Step 3: PredictIO Signature Validation {.tabset .tabset-pills}

Check which published IO signatures overlap with our candidate genes. This is **validation** — confirming that known signatures from other studies align with our findings.

```{r Check overlap between PredictIO signatures}
sig_candidate_overlap <- list()

for (file in rda_files) {
  sig_obj <- get(load(file))
  sig_name <- if ("signature_name" %in% colnames(sig_obj) && length(unique(sig_obj$signature_name)) > 0)
    unique(sig_obj$signature_name)[1]
  else
    tools::file_path_sans_ext(basename(file))

  if (!"gene_name" %in% colnames(sig_obj)) next

  genes <- sig_obj$gene_name
  matched <- intersect(candidate_genes, genes)

  if (length(matched) > 0) {
    sig_candidate_overlap[[sig_name]] <- data.frame(
      signature = sig_name,
      n_sig_genes = length(genes),
      n_candidate_overlap = length(matched),
      overlapping_genes = paste(sort(matched), collapse = ", ")
    )
  }
}

sig_overlap_df <- do.call(rbind, sig_candidate_overlap)
sig_overlap_df <- sig_overlap_df[order(-sig_overlap_df$n_candidate_overlap), ]

kable(sig_overlap_df, caption = "PredictIO Signatures Overlapping with Candidate Genes") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE) %>%
  scroll_box(height = "400px")
```

Test if the overlapping subset of each PredictIO signature predicts response.

> **Note:** These scores use only the genes that overlap with our candidates, not the full signature. Full signature scoring is done in Section 9 using the PredictioR package.

```{r predictio overlap response}
sig_score_results <- list()

for (sig_name in names(sig_candidate_overlap)) {
  overlap_genes <- strsplit(sig_candidate_overlap[[sig_name]]$overlapping_genes, ", ")[[1]]
  overlap_genes <- overlap_genes[overlap_genes %in% rownames(expr_data)]
  
  if (length(overlap_genes) < 2) next
  
  for (treat in sub_treat) {
    clin_sub <- clin_data[clin_data$treatment == treat, ]
    ids <- clin_sub$patientid[clin_sub$patientid %in% colnames(expr_data)]
    
    sig_score <- colMeans(expr_data[overlap_genes, ids], na.rm = TRUE)
    response <- clin_sub$response[clin_sub$patientid %in% ids]
    
    wt <- wilcox.test(sig_score ~ response)
    
    sig_score_results[[paste(treat, sig_name)]] <- data.frame(
      Treatment = treat,
      Signature = sig_name,
      n_genes = length(overlap_genes),
      p_value = wt$p.value,
      mean_R = mean(sig_score[response == "R"], na.rm = TRUE),
      mean_NR = mean(sig_score[response == "NR"], na.rm = TRUE)
    )
  }
}

sig_score_df <- do.call(rbind, sig_score_results)
sig_score_df$FDR <- p.adjust(sig_score_df$p_value, method = "BH")
sig_score_df$Direction <- ifelse(sig_score_df$mean_R > sig_score_df$mean_NR, "Up in R", "Up in NR")
sig_score_df <- sig_score_df[order(sig_score_df$p_value), ]

kable(sig_score_df, caption = "PredictIO Overlap Scores vs Response (Validation)") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE) %>%
  scroll_box(height = "400px")
```



```{r Boxplots predictio overlap, fig.width=10, fig.height=8}
# add gene names
wrap_genes <- function(genes, width = 45) {
  stringr::str_wrap(paste(genes, collapse = ", "), width = width)
}

# helper: wrap long gene lists nicely
wrap_genes <- function(genes, width = 45) {
  stringr::str_wrap(paste(genes, collapse = ", "), width = width)
}

for (treat in sub_treat) {

  top_sigs_treat <- sig_score_df %>%
    filter(Treatment == treat, FDR < 0.05) %>%
    arrange(FDR) %>%
    head(6)

  if (nrow(top_sigs_treat) == 0) next

  clin_sub <- clin_data[clin_data$treatment == treat, ]
  ids <- clin_sub$patientid[clin_sub$patientid %in% colnames(expr_data)]
  if (length(ids) < 2) next

  plot_list <- list()

  for (i in seq_len(nrow(top_sigs_treat))) {
    sig_name <- top_sigs_treat$Signature[i]
    fdr_val <- top_sigs_treat$FDR[i]

    n_total <- sig_candidate_overlap[[sig_name]]$n_sig_genes

    overlap_genes <- strsplit(sig_candidate_overlap[[sig_name]]$overlapping_genes, ",\\s*")[[1]]
    overlap_genes <- overlap_genes[overlap_genes %in% rownames(expr_data)]
    n_used <- length(overlap_genes)

    if (n_used < 2) next

    df <- data.frame(
      Score = colMeans(expr_data[overlap_genes, ids, drop = FALSE], na.rm = TRUE),
      Response = clin_sub$response[match(ids, clin_sub$patientid)]
    )

    genes_txt <- wrap_genes(overlap_genes, width = 45)

    plot_list[[sig_name]] <- ggplot(df, aes(x = Response, y = Score, fill = Response)) +
      geom_boxplot(outlier.shape = NA) +
      geom_jitter(width = 0.2, alpha = 0.5, size = 1) +
      scale_fill_manual(values = c("R" = "#E74C3C", "NR" = "#3498DB")) +
      labs(
        title = sig_name,
        subtitle = paste0(
          "FDR = ", formatC(fdr_val, format = "e", digits = 2),
          "  (", n_used, "/", n_total, " genes)\n",
          genes_txt
        ),
        y = "Overlap score (mean expression)",
        x = NULL
      ) +
      theme_minimal() +
      theme(
        legend.position = "none",
        plot.title = element_text(size = 9),
        plot.subtitle = element_text(size = 6, lineheight = 0.95)
      )
  }

  if (length(plot_list) == 0) next

  combined <- wrap_plots(plot_list, ncol = 3) +
    plot_annotation(title = paste("PredictIO Overlap Scores -", treat))

  print(combined)
}
```

#### Step 4: Heatmaps {.tabset .tabset-pills}

Heatmap of the top 20 significant genes (FDR < 0.05) from gene-response correlation, clustered by expression pattern (ward.D2). Columns are patients, rows are genes, expression is z-score scaled.

**Column annotations:**

- **Response**: pathological complete response (R) vs residual disease (NR)
- **Receptor**: hormone receptor and HER2 status (TN, HR+HER2-, HR+HER2+, HR-HER2+)

```{r Heatmap gene-response, fig.width=18, fig.height=7}
for (treat in sub_treat) {

  # -------------------------
  # 1) ALL significant genes
  # -------------------------
  sig_genes <- gene_response_df %>%
    filter(Treatment == treat, FDR < 0.05) %>%
    arrange(FDR) %>%
    pull(Gene) %>%
    unique()

  if (length(sig_genes) < 2) next

  top20_genes <- head(sig_genes, 20)

  clin_sub <- clin_data[clin_data$treatment == treat, ]
  ids <- clin_sub$patientid[clin_sub$patientid %in% colnames(expr_data)]
  if (length(ids) < 2) next

  # -------------------------
  # Annotations (match order)
  # -------------------------
  annot_col <- data.frame(
    Response = clin_sub$response[match(ids, clin_sub$patientid)],
    row.names = ids
  )

  annot_colors <- list(
    Response = c("R" = "#2ECC71", "NR" = "#F39C12")
  )

  if ("Receptor.Subtype" %in% colnames(clin_sub)) {
    vals <- clin_sub$Receptor.Subtype[match(ids, clin_sub$patientid)]
    if (sum(!is.na(vals)) > 0) {
      annot_col$Receptor <- vals
      annot_colors$Receptor <- c(
        "TN" = "#D62728", "HR+HER2-" = "#1F77B4",
        "HR+HER2+" = "#9467BD", "HR-HER2+" = "#E377C2"
      )[unique(na.omit(vals))]
    }
  }

  # -------------------------
  # Matrices
  # -------------------------
  mat_all <- expr_data[sig_genes, ids, drop = FALSE]
  mat_top20 <- expr_data[top20_genes, ids, drop = FALSE]

  # -------------------------
  # Heatmap A: ALL genes
  # -------------------------
  pheatmap(mat_all,
    annotation_col = annot_col,
    annotation_colors = annot_colors,
    scale = "row",
    show_colnames = FALSE,
    show_rownames = FALSE,
    main = paste0("Significant genes (FDR < 0.05) - ", treat, " (n=", length(sig_genes), ")"),
    clustering_method = "ward.D2",
    color = colorRampPalette(c("#2166AC", "white", "#B2182B"))(100),
    border_color = NA,
    fontsize = 10,
    treeheight_col = 30,
    treeheight_row = 30,
    cutree_cols = 2
  )

  # -------------------------
  # Heatmap B: TOP 20 genes
  # -------------------------
  pheatmap(mat_top20,
    annotation_col = annot_col,
    annotation_colors = annot_colors,
    scale = "row",
    show_colnames = FALSE,
    show_rownames = TRUE,
    fontsize_row = 9,
    main = paste0("Top 20 genes (by FDR) - ", treat),
    clustering_method = "ward.D2",
    color = colorRampPalette(c("#2166AC", "white", "#B2182B"))(100),
    border_color = NA,
    fontsize = 10,
    treeheight_col = 30,
    treeheight_row = 30,
    cutree_cols = 2
  )
}
```



### 9. Full PredictIO Signature Scoring

* Score each patient using the **PredictioR** R package ([GitHub](https://github.com/bhklab/PredictioR))
* Uses the package's built-in scoring function rather than simple mean expression
* Stratify by treatment arm to detect treatment-specific associations
* These signature scores become genomic features for radiogenomic integration

* Signature metadata (e.g., PMID, method, cancer context) is in:
  [https://github.com/bhklab/SignatureSets/blob/main/data-raw/signature_information_io.csv](https://github.com/bhklab/SignatureSets/blob/main/data-raw/signature_information_io.csv)
* Note: current IO signatures are pan-cancer / not breast-specific (**#TODO: add breast-specific signatures if needed**).

```{r Stratified signature analysis}
# TODO: Implement PredictioR-based signature scoring per treatment arm
# TODO: Confirm/control group mapping (e.g., Paclitaxel arms as control)
# TODO: PredictIO signatures are IO-focused and largely pan-cancer
```

---

## Section 3: Radiomics Analysis

Process MRI-derived radiomic features to identify imaging biomarkers associated with treatment response, and prepare features for radiogenomic integration.

### 1. Load & Clean Radiomic Features
Check how many patients, how many features

```{r load radiomics features}
# Load radiomics features
ISPY_radiomics  <- read.csv(file.path(here(), "data/results/radiomics/TCIA_ISPY2_feature_extracted_v1.csv"))
ACRIN_radiomics <- read.csv(file.path(here(), "data/results/radiomics/TCIA_ACRIN-6698_feature_extracted.csv"))
# missing_radiomics <- read.csv(file.path(here(), "data/results/radiomics/missing_samples.csv"))

# Combine radiomics
full_radiomics <- bind_rows(ISPY_radiomics, ACRIN_radiomics)

# Check for all-NA rows
feature_cols <- grep("^original", colnames(full_radiomics), value = TRUE)
all_na_rows <- rowSums(!is.na(full_radiomics[, feature_cols])) == 0
cat("All-NA feature rows:", sum(all_na_rows), "\n")

# Load index files
ispy_index  <- read.csv(file.path(here(), "data/procdata/radiomics/ISPY2/mit_ISPY2_index.csv"))
acrin_index <- read.csv(file.path(here(), "data/procdata/radiomics/ISPY2/mit_ACRIN-6698_index.csv"))
common_cols <- intersect(names(ispy_index), names(acrin_index))
for (col in common_cols) {
  if (class(ispy_index[[col]]) != class(acrin_index[[col]])) {
    ispy_index[[col]] <- as.character(ispy_index[[col]])
    acrin_index[[col]] <- as.character(acrin_index[[col]])
  }
}
full_index <- bind_rows(ispy_index, acrin_index)

# Extract SampleID from filepath
full_index$SampleID <- str_extract(full_index$filepath, "^[^/]+")
tp_lookup <- full_index %>%
  filter(Modality == "MR") %>%
  select(SampleID, PatientID, AcquisitionDate) %>%
  distinct() %>%
  group_by(PatientID) %>%
  arrange(AcquisitionDate) %>%
  mutate(TIME_POINT = paste0("T", row_number() - 1)) %>%
  ungroup()

# Join TIME_POINT to radiomics
full_radiomics <- full_radiomics %>%
  left_join(tp_lookup %>% select(SampleID, PatientID, TIME_POINT),
            by = c("patientid" = "SampleID"))

# Add patient_base for clinical merge
full_radiomics$patient_base <- gsub("^ISPY2-|^ACRIN-6698-", "", gsub("_.*", "", full_radiomics$patientid))

# Check
cat("TIME_POINT distribution:\n"); print(table(full_radiomics$TIME_POINT, useNA = "ifany"))
radiogenomic_patients <- intersect(full_radiomics$patient_base, merged_clin$patientid)
cat("Radiogenomic cohort:", length(radiogenomic_patients), "patients\n")

# Keep only overlapping patients
full_radiomics <- full_radiomics %>% filter(patient_base %in% radiogenomic_patients)
merged_clin <- merged_clin %>% filter(patientid %in% radiogenomic_patients)

kable(head(full_radiomics[, c("patientid", "patient_base", "PatientID", "TIME_POINT", feature_cols[1:5])], 10),
      caption = "Radiomic Features") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE) %>%
  scroll_box(height = "400px", width = "100%")

kable(head(merged_clin, 10), caption = "Clinical Data") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE) %>%
  scroll_box(height = "400px", width = "100%")
```


### 2. Select T0 (Baseline)

Filter to earliest timepoint per patient
This is your baseline radiomic profile

- **T0 baseline** for all patients
- **Delta-radiomics** for patients with T0 + T1
- **Volume change** visualization by response
- **Features correlated with volume** (expected, less interesting)
- **Features NOT correlated with volume** (interesting — correlate these with genomics next)

```{r assign timepoints}
# Samples per patient
radiomics_feature <- full_radiomics
samples_per_pt <- table(radiomics_feature$patient_base)
cat("Samples per patient:\n"); print(table(samples_per_pt))

# Verify existing timepoints
cat("TIME_POINT distribution:\n")
table(radiomics_feature$TIME_POINT)

# Select T0 (baseline)
radiomics_T0 <- radiomics_feature %>% filter(TIME_POINT == "T0")

# Patients with T0 + T1 (for delta-radiomics)
pts_T0 <- unique(radiomics_feature$patient_base[radiomics_feature$TIME_POINT == "T0"])
pts_T1 <- unique(radiomics_feature$patient_base[radiomics_feature$TIME_POINT == "T1"])
cat("Patients with T0 + T1 (for delta-radiomics):", length(intersect(pts_T0, pts_T1)), "\n")
```

### 3. Delta-Radiomics (T1 − T0)

delta = T1 value - T0 value

```{r delta radiomics}
# Select patients with both T0 and T1
radiomics_T1 <- radiomics_feature %>% filter(TIME_POINT == "T1")

delta_patients <- intersect(radiomics_T0$patient_base, radiomics_T1$patient_base)
cat("Patients with both T0 and T1:", length(delta_patients), "\n")

# Calculate T1 - T0 for each feature
T0_mat <- radiomics_T0 %>% filter(patient_base %in% delta_patients) %>% arrange(patient_base)
T1_mat <- radiomics_T1 %>% filter(patient_base %in% delta_patients) %>% arrange(patient_base)

delta_radiomics <- T1_mat[, feature_cols] - T0_mat[, feature_cols]
delta_radiomics$patient_base <- T0_mat$patient_base
cat("Volume change - Min:", min(delta_radiomics$volume_change), 
    "| Max:", max(delta_radiomics$volume_change),
    "| Mean:", round(mean(delta_radiomics$volume_change), 1),
    "| SD:", round(sd(delta_radiomics$volume_change), 1), "\n") 

cat("Delta-radiomics matrix:", nrow(delta_radiomics), "patients x", length(feature_cols), "features\n")

DT::datatable(delta_radiomics)
```

Volume change alone (T1 − T0) is not a strong predictor of response this early

```{r visualize volume change}
# Visualize change in volume (VoxelVolume)
delta_radiomics$volume_change <- delta_radiomics$original_shape_VoxelVolume

ggplot(delta_radiomics, aes(x = volume_change)) +
  geom_histogram(bins = 30, fill = "#3498DB", color = "white") +
  geom_vline(xintercept = 0, linetype = "dashed", color = "red") +
  labs(title = "Distribution of Tumor Volume Change (T1 - T0)",
       x = "Volume Change (voxels)", y = "Count") +
  theme_minimal()

# Volume change by response
delta_with_clin <- merge(delta_radiomics, merged_clin[, c("patientid", "response", "treatment")],
                          by.x = "patient_base", by.y = "patientid")

ggplot(delta_with_clin, aes(x = response, y = volume_change, fill = response)) +
  geom_boxplot(outlier.shape = NA) +
  geom_jitter(width = 0.2, alpha = 0.3, size = 1) +
  scale_fill_manual(values = c("R" = "#2ECC71", "NR" = "#F39C12")) +
  labs(title = "Tumor Volume Change (T1 - T0) by Response",
       x = "Response", y = "Volume Change") +
  theme_minimal() +
  theme(legend.position = "none")

# Check extreme outliers
vol_threshold <- mean(delta_radiomics$volume_change) + 3 * sd(delta_radiomics$volume_change)
cat("Outlier threshold (3 SD):", round(vol_threshold), "\n")
outliers <- delta_with_clin %>% filter(abs(volume_change) > abs(vol_threshold))
cat("Outliers:", nrow(outliers), "\n")
print(outliers[, c("patient_base", "volume_change", "response")])
```

```{r features correlated with volume, fig.width=12, fig.height=13}
# Correlate each feature change with volume change
vol_cor <- sapply(feature_cols, function(f) {
  cor(delta_radiomics[[f]], delta_radiomics$volume_change, use = "complete.obs")
})

vol_cor_df <- data.frame(
  Feature = names(vol_cor),
  Correlation = vol_cor,
  Abs_Correlation = abs(vol_cor)
) %>% arrange(desc(Abs_Correlation))

# Features highly correlated with volume (|r| > 0.7)
cat("Features highly correlated with volume (|r| > 0.7):", sum(vol_cor_df$Abs_Correlation > 0.7), "\n")

# Features NOT correlated with volume (|r| < 0.3) - interesting for genomics
cat("Features independent of volume (|r| < 0.3):", sum(vol_cor_df$Abs_Correlation < 0.3), "\n")

ggplot(vol_cor_df, aes(x = reorder(Feature, Abs_Correlation), y = Correlation)) +
  geom_col(fill = "#3498DB") +
  coord_flip() +
  labs(title = "Features Correlated with Volume Change",
       x = NULL, y = "Pearson Correlation with Volume Change") +
  theme_minimal()

# Plot top 20 correlated features
ggplot(head(vol_cor_df, 20), aes(x = reorder(Feature, Abs_Correlation), y = Correlation)) +
  geom_col(fill = "#3498DB") +
  coord_flip() +
  labs(title = "Top 20 Features Correlated with Volume Change",
       x = NULL, y = "Pearson Correlation with Volume Change") +
  theme_minimal()

# Plot features NOT correlated with volume
independent_features <- vol_cor_df %>% filter(Abs_Correlation < 0.3)
cat("\nVolume-independent features (candidates for genomic correlation):\n")
```



```{r}
sessionInfo()
```

3. Quality Check

Filter out small/failed segmentations
Check for missing values
Check how many patients overlap with genomics (982)

4. Delta-Radiomics (T1 − T0)

Filter patients with both T0 and T1
Calculate T1 minus T0 for each feature
Visualize volume change
Identify features correlated with volume
Identify features NOT correlated with volume (interesting ones)

5. Correlate Radiomics with Response

Test each radiomic feature vs R/NR per treatment
Similar to gene-response correlation in Section 8

6. READII (later)

Negative control images
Extract same features

