---
title: "I-SPY Project Data Processing"
author: "Nasim Bondar Sahebi and Katy Scott"
date: "`r format(Sys.time(), '%Y-%m-%d')`"
output:
  rmdformats::readthedown:
    toc_depth: 3
    theme: lumen
    highlight: espresso
    code_folding: show
  rmdformats: default
affiliation: Princess Margaret Cancer Center, University Health Network, Toronto,
  Canada
params:
  palette: npg
  size: 1
  showlabels: false
  ellipse: true
css: styles/custom.css
link-citations: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
knitr::opts_knit$set(root.dir = normalizePath(".."))
```
---
## Overview

This Script **processes and prepares** the **I-SPY2 dataset**, focusing on:
- **Cleaning, merging, and structuring** genomics and radiomics clinical data.
- **Identifying common patients** between datasets.
- **Ensuring MRI segmentation consistency** for further analysis.

### Dataset Summary
- **Study Reference**: [PMID: 35623341](https://pubmed.ncbi.nlm.nih.gov/35623341/)  
- **Clinical Trial**: [ISPY2/NCT01042379](https://clinicaltrials.gov/study/NCT01042379)  
- **Data Source**: [GEO: GSE194040](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE194040)  
- **Study Population**:  
  - **Genomics Data**: 987 patients (mRNA & RPPA profiling)  
  - **Radiomics Data**: 985 patients (MRI scans)  
- **Platforms Used**:
  - **Genomics**: Agilent 44K arrays  
  - **Radiomics**: MRI (DCE-MRI)   

### Patient and Treatment Details
- **Total Patients (RNA-seq)**: 987  
- **Total Patients (MRI)**: 985 (719 I-SPY2 cases, 266 ACRIN-6698 cases)  
- **Treatment Groups**:
  - **IO + Chemo**: 69 patients  
  - **Chemo + Targeted Therapy**: 687 patients  
  - **Chemo Alone**: 179 patients  
  - **Targeted Therapy Alone**: 52 patients  
- **Key Treatment Arms**:
  - Paclitaxel ± Trastuzumab  
  - Veliparib + Carboplatin  
  - AMG-386 (Trebananib) ± Trastuzumab  
  - Neratinib, Ganetespib, Ganitumab, Pembrolizumab, Pertuzumab, T-DM1  
  
#### Treatment Arms
- **Chemotherapy ± IO**  
  - Paclitaxel ± Trastuzumab  
  - Veliparib + Carboplatin  
- **Targeted Therapies**  
  - AMG-386 (Trebananib) ± Trastuzumab  
  - Neratinib, Ganetespib, Ganitumab, Pembrolizumab, Pertuzumab, T-DM1
  
### Additional Information
- **Batch Correction**:  
  - **ComBat** was applied to correct batch effects in gene expression data.  
  - Adjusted ~800 patients; linear adjustments applied to remaining cases.  

- **Radiomics Data**:  
  - MRI dataset includes **4 serial DCE-MRI scans per patient**.  
  - Performed **before and during neoadjuvant chemotherapy** to track tumor volume changes and response.  

- **Data Availability**:  
  - 📂 **Genomics**: Available in [GEO: GSE194040](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE194040).  
  - 📂 **Imaging**: MRI datasets accessible via [The Cancer Imaging Archive (ISPY2 Collection)](https://www.cancerimagingarchive.net/collection/ispy2/).  

- **Key Reference**:  
  - Wolf DM et al., *Cancer Cell*, 2022. [PMID: 35623341](https://pubmed.ncbi.nlm.nih.gov/35623341).  

---

## Section 1: Data Processing Steps

### Load Required Libraries
 
```{r libraries,include=TRUE,message=FALSE,warning=FALSE}
library(here)
library(MultiAssayExperiment) 
library(knitr)
library(readxl)
library(ggplot2)
library(dplyr)
library(tidyr)
library(ggvenn)
library(kableExtra) 
```

### Load & Extract Data {.tabset .tabset-pills}

Load MultiAssayExperiment `.rds` File**
This step extracts **clinical, expression(Microarray quantile normalizated), and annotation data** for further analysis.

```{r load multiassay and extract Data}
# Define base directory.
input_dir <- file.path(here(), "data/rawdata/genomics")
output_dir <- file.path(here(), "results/genomics")

# Load dataset
mae <- readRDS(file.path(input_dir, "ICB_Wolf.rds"))
mae <- readRDS("../..data/rawdata/genomis/ICB_Wolf.rds")
# Extract clinical, expression, and annotation data
clin <- data.frame(colData(mae))  
expr <- assays(mae)[["expr"]] # dim 18348 x 987
annot <- data.frame(rowData(mae@ExperimentList$expr))

# Create a named vector to map gene_id to gene_name
gene_map <- setNames(annot$gene_name, annot$gene_id)
rownames(expr) <- gene_map[rownames(expr)]
rownames(annot) <- annot$gene_name
```

#### Clinical Data 

Display first 50 rows of  Clinical Data.

```{r clin summery}
# Display clin 
kable(head(clin,50), caption = "Genomics Clinical Data") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE) %>%
  scroll_box(height = "400px")
```

#### Microarray Expression Data 

Display first 20 rows of  quantile-normalized Microarray data

```{r expr summery}
# Display the first 20 rows 
kable(head(expr, 20), caption = "Expression Data-Microarray") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE) %>%
  scroll_box(height = "400px")
```

#### Annotation Data 

Displays gene annotation for the dataset.

```{r data summery}
# Display the first 20 rows 
kable(head(annot, 20), caption = "Gene Annotation Data") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE) %>%
  scroll_box(height = "400px")
```

### Match Patients: Genomics & Radiomics Integration

This section identifies overlapping patients between genomics (gene expression) and radiomics (MRI scans) datasets.

#### Step 1: Load Radiomics Metadata

The radiomics dataset contains dynamic contrast-enhanced MRI **(DCE-MRI)** scans used to monitor tumor response to treatment.  

**Metadata Source:**  
- **File:** `ISPY2-Imaging-Cohort-1-Clinical-Data.xlsx`  
- **Access:** [The Cancer Imaging Archive - I-SPY2 Collection](https://www.cancerimagingarchive.net/collection/ispy2/)  
- **Total Patients:** **985**  
  - **719** I-SPY2 cases  
  - **266** [ACRIN-6698 cases](https://www.cancerimagingarchive.net/collection/ispy1/)  

```{r Load Radiomics Metadata}
# Load radiomics clinical data
Radiomics_clin <- read_excel(file.path(input_dir, "ISPY2-Imaging-Cohort-1-Clinical-Data.xlsx"))  

# Display radiomics metadata with scrolling
kable(Radiomics_clin, caption = "Radiomics Clinical Data") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE) %>%
  scroll_box(height = "400px")
```

#### Step 2: Summary of Radiomics Data

- **Each patient has 4 MRI scans (T1-T4)**
- **MRI features include:**
  - Functional Tumor Volume (FTV)
  - Longest Tumor Diameter
  - Sphericity  
- **Data collected from 22+ clinical centers**   

#### Step 3: Identify Common Patients

To integrate genomics and radiomics data, we **match patient IDs** across datasets.

```{r check Patient Matching}
# Prepare genomic clinical data
Genomics_clin <- clin  
Genomics_clin$patientid <- sub("^X", "", Genomics_clin$patientid)  # Remove "X" prefix
rownames(Genomics_clin) <- Genomics_clin$patientid
colnames(expr) <- sub("^X", "", colnames(expr))

# Identify common patients
common_patients <- intersect(Genomics_clin$patientid, Radiomics_clin$Patient_ID)  

# Print count of common patients
cat("Patients with both microarray and MRI data:", length(common_patients), "\n")
```

#### Step 4: Visualize Patient Overlap

A Venn diagram illustrates genomics vs
. radiomics patient overlap.

```{r visualize patient overlap}
# Create Venn diagram
venn_data <- list(Genomics = Genomics_clin$patientid, Radiomics = Radiomics_clin$Patient_ID)

ggvenn(venn_data, fill_color = c("#A6CEE3", "#FB9A99"), stroke_size = 0.5, set_name_size = 4) +
  ggtitle("Overlap of Genomics (Microarray) and Radiomics (MRI) Patient IDs") +
  theme(plot.title = element_text(hjust = 0.5, size = 12, face = "bold"), plot.background = element_rect(fill = "white"), panel.background = element_blank())
```

## Merge Genomics & Radiomics Data

Merging both datasets based on **common patient IDs**.

**Adjustments:**  
- **Prefix MRI columns with `"Radiomics_"`**  
- **Ensure consistent patient IDs**  
- **Save merged dataset for analysis**  
 
```{r Merge}
# Add prefix to radiomics columns
colnames(Radiomics_clin)[-which(colnames(Radiomics_clin) == "Patient_ID")] <- paste0("Radiomics_", colnames(Radiomics_clin)[-which(colnames(Radiomics_clin) == "Patient_ID")])

# Merge datasets
merged_clin <- merge(Genomics_clin, Radiomics_clin, by.x = "patientid", by.y = "Patient_ID")

# Sort by patient ID
merged_clin <- merged_clin[order(merged_clin$patientid), ]

# Display merged dataset with scrolling
kable(head(merged_clin,50), caption = "Merged Clinical Data (Genomics & Radiomics)") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE) %>%
  scroll_box(height = "400px")

# Save dataset as CSV
data/results/genomics
write.csv(merged_clin, file = file.path(output_dir, "combined_clin_radiogenomic.csv"), row.names = FALSE)

```

### Segmentation Consistency Check

To ensure accurate **Functional Tumor Volume (FTV)** analysis in the I-SPY clinical trials, we verify that each patient has the expected **4 segmentation masks** from **Dynamic Contrast Enhanced (DCE) MRI** scans.  

### Segmentation Process in DCE-MRI

Each scan undergoes multiple segmentation steps:  
1. **Manual VOI (Volume of Interest):** A rectangular region around the enhancing tumor.  
2. **Background Masking:** Removes noise and fat saturation artifacts.  
3. **Percent Enhancement (PE) Threshold:** Filters non-enhancing tissue.  
4. **Connectivity Filter:** Ensures 3D tumor voxel connectivity.  
5. **OMIT Regions:** Manual exclusion of non-tumor areas.  

**Segmentation masks are stored as 2D DICOM images** aligned with the original **DCE MRI scans**.  
note: **Download masks from UCSF Box:** [Link](https://ucsf.box.com/s/d09xetaatzgnz2p9bxlqh50rrzilgcoj)  


### Segmentation Metadata {.tabset .tabset-pills}

#### ISPY1: MRI Series Metadata  
**Metadata for ISPY1** is stored in `ispy1_dce_series_info.xlsx`, containing data for **221 patients**. This dataset includes MRI series metadata for each patient, covering:  

- **Patient ID & Time Point (T1–T4)**  
- **MRI Series Description & Unique ID**  
- **Signal Enhancement Ratio (SER) Calculation Indicator**  
- **OMIT Region Counts (Non-Tumor Exclusion)**   

```{r ISPY1 segmentation metadata}
# Load ISPY1 segmentation metadata
ispy1_dce_series_info <- read_excel(
  path = file.path(input_dir, "ispy1_dce_series_info.xlsx"), 
  sheet = "DCE Series, All 20160823"
) 

# Display ISPY1 segmentation metadata
kable(head(ispy1_dce_series_info, 20), caption = "DCE MRI Series Information") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE) %>%
  scroll_box(height = "400px")
```

#### ISPY2: MRI Series Metadata  

Metadata for ISPY2 was extracted from `series-data1741883133411.csv`, containing data for **719 patients**. This dataset includes MRI series metadata for each patient, covering:  

- **Subject ID**  
- **Study Description** (e.g., `"ISPY2_MRI_T0"`)  
- **Number of Images**  
- **Modality** (e.g., `"MR"`, `"SEG"`)  

```{r segmentation metadata}
# Load ISPY2 segmentation metadata
ispy2_dce_series_info <- read.csv(file.path(input_dir, "series-data1741883133411.csv"))

# Rename the "ispy2_dce_series_info" column to "PATIENT_ID"
colnames(ispy2_dce_series_info)[colnames(ispy2_dce_series_info) == "Subject.ID" ] <- "PATIENT_ID"

# Extract time points from the "Study.Description" column
ispy2_dce_series_info$TIME_POINT <- gsub(".*(T[0-9]+).*", "\\1", ispy2_dce_series_info$Study.Description)

# Display ISPY2 segmentation metadata
kable(head(ispy2_dce_series_info, 20), caption = "ISPY2 DCE MRI Series Information") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE) %>%
  scroll_box(height = "400px")

# Select only PATIENT_ID and TIME_POINT columns from both datasets
ispy1_selected <- ispy1_dce_series_info[, c("PATIENT_ID", "TIME_POINT")]
ispy2_selected <- ispy2_dce_series_info[, c("PATIENT_ID", "TIME_POINT")]

# Merge based on PATIENT_ID and TIME_POINT
ispy1and2 <- merge(ispy1_selected, ispy2_selected, by = c("PATIENT_ID", "TIME_POINT"), all = TRUE)
```

### Check Segmentation Consistency Across Time Points (T1-T4)  

Each patient (Total: 940) should have at least four segmentation masks, corresponding to their four MRI scans.

```{r Check Segmentation}
# Count segmentations per patient for each time point
series_per_patient <- ispy1and2 %>%
  group_by(PATIENT_ID, TIME_POINT) %>%
  summarize(Series_Count = n(), .groups = 'drop') %>%
  pivot_wider(names_from = TIME_POINT, values_from = Series_Count, values_fill = 0) %>%
  mutate(across(starts_with("T"), as.numeric),
         Total_Segmentations = rowSums(select(., starts_with("T")), na.rm = TRUE))

# Display segmentation counts per patient
kable(series_per_patient, caption = "Segmentation Counts per Patient") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE) %>%
  scroll_box(height = "400px")
```

### Patients with Missing Segmentations

Patients missing **one or more segmentation masks** need further review (24 patients).

```{r missing segmentations}
# Identify patients with <4 segmentations
patients_less_than_4 <- series_per_patient %>%
  filter(Total_Segmentations < 4) %>%
  select(PATIENT_ID, Total_Segmentations)

# Display patients missing segmentations
kable(patients_less_than_4, caption = "Patients with Less Than 4 Segmentation Masks") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE) %>%
  scroll_box(height = "400px")
```

---

## Section 2:  Genomic Analaysis 

### Step-by-Step Plan
1. Subset to Radiogenomic Cohort
Subset both clinical and genomic (microarray) data to the common set of 982 patients with both genomic and radiomic data.

2. Expression Filtering
Remove genes with low or zero expression across most samples (e.g., keep genes expressed in >10% of samples).
Optional: Apply log2 transformation if needed.

##############
Preparing expression data (RNA-seq) for analysis:

1. Restrict expression data to protein-coding genes.
2. Remove low/zero expression genes.

Total 16,256 genes remain for downstream analyses. 

```{r feature/gene reduction}

# Filter out protein coding genes
# Remove low/0 expression
# Filter out genes with low variance
# See what number of features there are at that point
# Maximum ~15k
# Correlation of gene features with treatment response
# Pathway analysis(maybe)

# 1. Subset to common patients
expr_data <- expr[, common_patients]
annot_data <- annot[annot$gene_name %in% common_patients , ]
clin_data <- merged_clin[merged_clin$patientid %in% common_patients, ]

# 2. Restrict to Protein-Coding Genes.
annot_proteincoding <- annot[annot$gene_type == "protein_coding",] # 17547 protein coding genes.
expr_data <- expr_data[rownames(expr_data) %in% rownames(annot_proteincoding),]

# 3. Filter Low/Zero Expression Genes]
#Microarray with normalization log2gNor means the data is log2 transformed and then quantile normalized.
# r <- as.numeric(apply(expr_data, 1, function(i) sum(i < 5)))
# remove <- which(r > dim(expr_data)[2] * 0.5) 

# Remove rows from the matrix
filtered_expr <- expr_data[rowMeans(expr_data) > log2(5), ]  #dim is 17547 x 982
DT::datatable(round(filtered_expr [1:8, 1:8], 3))


# 4. correlated gene features with treatment response using predioR pacakage ( logistic regressino fro binary PCR/Response)

# 5. Use limma to find differentially expressed genes (DEGs).
#significant_genes <- deg_results[deg_results$adj.P.Val < 0.05 & abs(deg_results$logFC) > 1, ]

# 7. Visualizations
# volcano plot to showing the most significant genes ( proetim cocindg - nonprotein coding genes)


# any other visluzations that is nice


# we can also use somepath way analasyiss to see the pathways the gene are working in 

```


